---
title: "Practicum Notebook"
output: html_notebook
author: Calvin Crosby, Kristine Umeh
email: crosby.ca@northeastern.edu, umeh.k@northeastern.edu
---
Required packages to be installed:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Connecting to the database 
```{r}
# 1. Library
library(RMySQL)
# 2. Settings
db_user <- 'admin'
db_password <- 'Swayze96'
db_name <- 'birddb'
db_host <- 'birddb.cd3w70rzjf0l.us-east-2.rds.amazonaws.com'
db_port <- 3306

# 3. Read data from db
mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)
```
Drop statements to empty database
```{sql connection=mydb}
DROP TABLE IF EXISTS incidents
```

```{sql connection=mydb}
DROP TABLE IF EXISTS airlines
```

```{sql connection=mydb}
DROP TABLE IF EXISTS airports
```
Schema defined for airlines table
```{sql connection=mydb}
create table airlines(
  aid INT NOT NULL,
  code VARCHAR(200)NOT NULL DEFAULT 'unknown',
  airline VARCHAR(200) UNIQUE NOT NULL DEFAULT 'unknown',
  PRIMARY KEY (aid)
);

```
Schema defined for airports table
```{sql connection=mydb}
create table airports(
pid INT NOT NULL,
code VARCHAR(200)NOT NULL DEFAULT 'unknown',
name VARCHAR(200)UNIQUE NOT NULL DEFAULT 'unknown',
city VARCHAR(200) NOT NULL DEFAULT 'unknown',
state VARCHAR(200)NOT NULL DEFAULT 'unknown',
country VARCHAR(200)NOT NULL DEFAULT 'unknown',
PRIMARY KEY (pid)
);
```
Schema defined for incidents table
```{sql connection=mydb}
create table incidents (
  iid INT NOT NULL,
  dateOnly DATE DEFAULT NULL,
  depPort INT,
  arrPort INT,
  airline INT,
  aircraft VARCHAR(200)NOT NULL DEFAULT 'unknown',
  flightPhase VARCHAR(12)NOT NULL DEFAULT 'unknown',
  impact VARCHAR(200) NOT NULL DEFAULT 'unknown',
  PRIMARY KEY (iid),
  FOREIGN KEY (airline) REFERENCES airlines(aid),
  FOREIGN KEY (depPort) REFERENCES airports(pid),
  FOREIGN KEY (arrPort) References airports(pid)
)
```

```{r}
install.packages("hash")
```
A hash map is made to map possible encountered values to harmonized values. It is used by
my harmonizer function
```{r}
library(hash)

flightMap <- hash()
flightMap[["Take-off run"]] <- "takeoff"
flightMap[["Landing Roll"]] <- "landing"
flightMap[["Climb"]] <- "inflight"
flightMap[["Approach"]] <- "inflight"
flightMap[["Descent"]] <- "inflight"
flightMap[["Taxi"]] <- "takeoff"
flightMap[["Parked"]]<- "unknown"
flightMap[[" "]]<- "unknown"
```
Defining my phase harmonizer function. Tests are run at the end to show functionality
```{r}
# This function is used to harmonize the flight phase data from the csv.
# It references the flightMap hashmap and returns the appropriate string corresponding
# to a known phase or it returns "unknown" if an unrecognized or NULL value is
# encountered.
phaseHarmonizer <- function(flightPhase){
  if (is.null(flightPhase)){
    return("unknown")
  }
  if (has.key(flightPhase,flightMap)){
    return(flightMap[[flightPhase]])
  }
  return("unknown")
}
phaseHarmonizer("Take-off run") 
phaseHarmonizer('Take-off run')
phaseHarmonizer("Climbing")
phaseHarmonizer("In the ocean lol")
phaseHarmonizer("Landing Roll")
phaseHarmonizer(NULL)
phaseHarmonizer("Taxi")
phaseHarmonizer("Parked")
```
This reads in the csv file and stores it in a data frame. Duplicates are removed
```{r}
require(readr)
library(tidyverse)
birds <- read_csv('BirdStrikesData.csv')
birds[!duplicated(birds$`Record ID`), ] #remove duplicates
```

This uses my custom phaseHarmonizer function to update the birds data frame flight Phase
column. It checks for additional cases like empty strings and null values
```{r}
nrows <- nrow(birds)

for (i in 1:nrows){
  if (is.na(birds[[i,14]]) | birds[[i,14]] == '')
    birds[[i,14]] == "unknown"
  else if (is.null(birds[[i,14]]))
    birds[[i,14]] == "unknown"
  else if (birds[[i,14]]== ""){
    birds[[i,14]] <- "unknown"
  }
  else{
    temp <- birds[[i,14]]
    temp <- phaseHarmonizer(temp)
    birds[[i,14]] <- temp
    }
  }
```

```{r}
birds
```

```{r}
library(anytime)

print(typeof(birds))
print(typeof(birds[,12]))
#airlineData is used to populate airline table, aid is auto-incremented
airlineData <- data.frame(airline = distinct(birds[,12])) #look up table for airlines
airlineData <- transmute(airlineData, airline=`Aircraft..Airline.Operator`,aid=1:n())

#airportsData is used to populate the airports table, pid is auto-incremented
airportsData <- data.frame(name = birds[,3],state=birds[,13])
airportsData <- distinct(airportsData, `Airport..Name`, .keep_all = TRUE)
airportsData <- transmute(airportsData, name = `Airport..Name`, state=`Origin.State`,pid=1:n())

airports.name <- data.frame(name = birds$`Airport: Name`)
airports.state <- data.frame(state = birds$`Origin State`)

#airPFull <-cbind(airports.name,airports.state)
#incidentsFull <-cbind(incidents.iid,incidents.date)


incidentsFull <- transmute(birds, iid = `Record ID`, dateOnly= anydate(`FlightDate`),airline=birds$`Aircraft: Airline/Operator`, aircraft=`Aircraft: Make/Model`, flightPhase=`When: Phase of flight`, name=birds$`Airport: Name`,impact=`Effect: Impact to flight`)

incidentsFull <- left_join(incidentsFull, airlineData, by = "airline")
incidentsFull <- left_join(incidentsFull, airportsData, by = "name")

# IncidentsFull is the dataframe for the incidents table
tail(incidentsFull)

```

```{sql connection=mydb}
DROP TABLE IF EXISTS airlinesAUX

```

```{sql connection=mydb}
DROP TABLE IF EXISTS airportsAUX

```

```{sql connection=mydb}
DROP TABLE IF EXISTS incidentsAUX
```
Writing data into our auxillary tables
```{r}
dbWriteTable(mydb,"incidentsAUX",incidentsFull,overwrite=F,append=T)
dbWriteTable(mydb,"airlinesAUX",airlineData,overwrite=F,append=T)
dbWriteTable(mydb,"airportsAUX",airportsData,overwrite=F,append=T)
```
Inserting data into the airlines auxillary table
```{sql connection=mydb}
INSERT INTO airlines(aid,airline) SELECT aid,airline FROM airlinesAUX;
```
Showing data in airlines table in the database
```{sql connection=mydb}
SELECT * FROM airlines
```
Inserting data into airports table from airport auxillary table
```{sql connection=mydb}
INSERT INTO airports(pid,name,state) SELECT pid,name,state FROM airportsAUX;
```
Showing the airports table in the database
```{sql connection=mydb}
SELECT * FROM airports
```
Inserting data into incidents table from the incidents auxillary table
```{sql connection=mydb}
INSERT INTO incidents(iid,dateOnly,depPort,arrPort,airline,aircraft,flightPhase,impact)
SELECT iid,dateOnly,pid,pid,aid,aircraft,flightPhase,impact
FROM incidentsAUX;
```
Taking a look at my incidents table in the database
```{sql connection=mydb}
SELECT * FROM incidents
```

Query 1
```{sql connection=mydb}
SELECT 
airlines.airline AS 'ARRIVING AIR LINE',
COUNT(incidents.iid)AS 'NUM OF INCIDENTS'
FROM incidents
INNER JOIN airports ON airports.pid = incidents.arrPort
INNER JOIN airlines ON airlines.aid = incidents.airline
WHERE
(
(airports.name like '%LaGuardia%')
AND
(incidents.flightPhase = 'landing')
)
GROUP BY airlines.airline
```

Query 2
```{sql connection=mydb}
SELECT * FROM (
SELECT depPort,arrPort,airports.pid,airports.name,
COUNT(iid) AS 'COUNT_OF_INCIDENTS'
FROM incidents
INNER JOIN airlines ON incidents.airline = airlines.aid
INNER JOIN airports ON incidents.depPort = airports.pid
WHERE
(
  (
    (
    UPPER(airlines.airline) NOT LIKE '%BUSINESS%'
    )
    AND
    (UPPER(airlines.airline) NOT LIKE '%MILITARY%'
    )
    AND
    (UPPER(airlines.airline)NOT LIKE '%CORP%'
    )
    AND
    (UPPER(airlines.airline)NOT LIKE '%EXEC%'
    )
    AND
    (UPPER(airlines.airline)NOT LIKE '%GOVERNMENT%'
    )
    AND
    (UPPER(airlines.airline)NOT LIKE '%PRIVATE%'
    )
    AND 
    (UPPER(airlines.airline)NOT LIKE '%UNKNOWN%'
    )
  )
)
GROUP BY depPort, arrPort,airports.pid,airports.name
ORDER BY COUNT_OF_INCIDENTS DESC
) AS COUNT_OF_COMMERCIAL_INC
LIMIT 1
```

Query 3
```{sql connection=mydb}
SELECT
YEAR(dateOnly) AS 'YEAR',
COUNT(*) AS 'BIRD_STRIKES_PER_YEAR'
FROM
incidents
GROUP BY YEAR(dateOnly)
ORDER BY YEAR(dateOnly) ASC
```
Plotting the Bird Incidents from 2005 to 2011
```{r}
library(ggplot2)
library(sqldf)
frame <- dbGetQuery(mydb,"SELECT
YEAR(dateOnly) AS 'YEAR',
COUNT(*) AS 'BIRD_STRIKES_PER_YEAR'
FROM
incidents
GROUP BY YEAR(dateOnly)
ORDER BY YEAR(dateOnly) ASC")

filtered <- subset(frame, YEAR>=2005 & YEAR<=2011)
filteredmore <-subset(filtered,!is.null(YEAR))
print(filteredmore)
year <- filtered[,1]
bird_strikes <- filtered[,2]
#create data frame (frame) with above query using sqldf and pass into this:
ggplot(filtered, aes(x=year,y=bird_strikes,group=1,colour="Bird Strikes"))+geom_line()+ geom_point()+
  labs(title= "Bird Strikes Per Year vs. Year Recorded", x = "Year",y= "Bird Strikes")+scale_color_discrete(name = "Legend", labels = c("Bird Strikes "))+ theme(plot.title = element_text(hjust = 0.5))

```

```{sql connection=mydb}
DROP PROCEDURE IF EXISTS delAirline;
```
Making our Stored Procedure
```{sql connection=mydb}
CREATE PROCEDURE delAirline(IN airlineid INT)  
BEGIN
  DELETE 
  FROM incidents
      WHERE incidents.airline = airlineid;
  DELETE FROM airlines
  WHERE airlines.aid = airlineid;
END;
```

--Calling our stored procedure to delete the airline with aid 20
```{sql connection=mydb}
CALL delAirline(20);
```
Proving the records have been removed
```{sql connection=mydb}

SELECT * FROM airlines
WHERE airlines.aid = 20;


```
Proving the records have been removed
```{sql connection=mydb}

SELECT * FROM incidents
WHERE incidents.airline = 20;
```
## Disconnect from Database

```{r}
dbDisconnect(mydb)
```

